Object detection for Verification Based Annotation

### 2019-09-09 paper.tex
Updates from Overleaf

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

\title{Object detection for Verification Based Annotation}
\author{\IEEEauthorblockN{Oliver Batchelor\IEEEauthorrefmark{1} and
Richard Green\IEEEauthorrefmark{2}}
\IEEEauthorblockA{Department of Computer Science,
University of Canterbury\\
Christchurch, New Zealand\\
Email: \IEEEauthorrefmark{1}oliver.batchelor@canterbury.ac.nz,
\IEEEauthorrefmark{2}richard.green@canterbury.ac.nz}}

@@

\newacronym{NMS}{NMS}{Non Maxima Suppression}

@@

In this paper we examine the properties of an object detector for the purpose of Verification Based Annotation, where annotation is performed by having machine annotations checked and verified (and corrected, in this work) by a human annotator. We present our approach to object detection, where with a few modifications from standard practice, an object detector \cite{Lin2017} can be robust, without a large degree of parameter tweaking, for training object detection incrementally in order to annotate a wide variety of small-scale datasets with a variety of image resolution and object size.

Our approach is to use high resolution images, training on image crops (as opposed to the usual practice of resizing input images to a fixed resolution), and find this method successful, being more accurate and robust than down-scaling. A particular interest is the effect of localisation noise and systematic bias in annotations. We characterise the impact on object detection performance and compare to human levels, we find noise has a large impact especially with fewer training examples.

@@

In this paper we examine the properties of an object detector used as part of a \gls{VBA} process. We are most interested in the robustness to varying parameters, such as image size, number and size of objects, and to initialisation. The ability to learn incrementally is key (and in particular, learn from a minimal number of examples).

@@

The object detector is based on a single-shot \gls{CNN} detector called RetinaNet \cite{Lin2017}, an anchor-box based single stage object detector. This detector was selected for its simplicity and efficiency, while having close to state-of-the-art accuracy. Some modifications from the method in \cite{Lin2017} and brief motivations for each are:

\begin{itemize}
    \item {\textbf {Sharing}}\par No shared weights between classification sub-networks at different pyramid levels, in order to make training much faster at the beginning.
    \item {\textbf {Non normalised loss}}\par 
    We do not normalise of the loss function (by the number of anchor matches), in order to accommodate images with no positive annotations.
    \item {\textbf {High resolution}}\par We train on crops of of high-resolution images, but for inference operate on complete images. The motivation is to make the object detection task as easy as possible; larger objects are easier to detect \cite{Lin2014}. 
    \item {\textbf {Cyclical learning rates}}\par We use of cyclical learning rates in training, to accommodate incremental additions to the training set.
    \item {\textbf{Extra decoder layers}}\par We used an extra residual blocks in order to combine shortcut connections after concatenation with features from lower resolutions.
\end{itemize}

@@

A selection of images of the datasets and typical annotations used in this paper can be seen in in Figure~\ref{fig:datasets_all}, and a summary of the different variations can be seen in Table~\ref{tab:resolutions}, most of the datasets were annotated by the author, a few annotated by third parties and one was annotated by both. The datasets used span a wide variety of image resolutions, and object sizes; the images in some cases are reduced resolution from the captured images, in other cases reducing the resolution would make object detection impossible.

@@

We looked at two different possibilities for performing inference on a full-resolution image using an object detector network trained only on crops of images: (a) pass in the full image using the property that the network is \emph{fully convolutional} and (b) tile images the same size as training crops and collapse the predictions using a combined \gls{NMS}. In order to facilitate this idea, the box annotations on the edge of images are estimates of the full bounds of the object, and the anchor boxes are not cropped to the edge of the image. In \cite{Duarte2010}, the approach of annotating the estimated full bounding box (the approach we have taken) was compared to annotating just the visible parts (as is more common practice) in a pedestrian detector. Estimating the full bounds was a slight improvement, however, the fusion of both was found to be better than either one.

The first method is to pass in the full image to the neural network, even though it has only been trained on much smaller crops. Object detection networks are flexible and work across a range of input image resolutions. All layers are either convolutions or do not reshape the feature maps (aside from up/down-sampling). As a result, passing in a larger image results in a larger feature map and set of box outputs at each layer of the pyramid. The concern is that the up/down-sampling behaviour is slightly different for input sizes depending on if they are odd or even, relative to powers of two. We found this to be a legitimate concern, though largely negated if the training crop size is set to a power of two or multiple of a power of two.

@@

For this experiment we vary the scale and crop size to test the idea that an high-resolution object detector can be trained on small parts of images, then the whole image used in inference.

@@

Using a larger crop size proved more accurate in all of the four datasets, and using a larger crop was beneficial in all cases, being more accurate and more stable to train. The larger size, however, trains much more slowly. In this case a factor of $3.3$ slower for half scale or half size crop, larger factors having less speedup due to constants such as loading the images. Despite the relatively low requirement for fast inference (to keep up with a single annotator); this could be used for example for image selection, to find images of most uncertainty for the purpose of active learning.

@@

When using the annotation tool, image annotations become available incrementally. Here we investigate the difference between training with all examples annotated up front compared to training with images added to the training set incrementally.

During annotation, the validation set is incrementally built. Here we test against the final validation sets. In future, it would be better to use cross-validation, in order to make better use of training data, and provide more accurate testing (at the expense of extra training time).

@@

A \gls{VBA} based system by nature of using an object detector trained on noisy inputs, will therefore not produce entirely accurate localised predictions. The hope is that such a system will eventually produce localisations that are \emph{good enough}, in order that predictions from the object detector can be accepted without taking up valuable annotator time. In this experiment, we aim to quantify how noise degrades performance, and therefore establish some idea of what guideline for the level of precision is required during annotation.

@@

Figure~\ref{fig:density_iou} shows in more detail the distribution of \gls{IOU} overlap for detections which have been modified. The density of transformed detections peaks at around $0.80$--$0.85$ depending on the dataset. It can be reasonably assumed that the real accuracy of the detections, which are used unmodified, would lie between that peak and $1.0$. Human annotation also has some degree of variability (reference \cite{Papadopoulos2017} gave a mean \gls{IOU} of 88 overlap for a box input method with Pascal VOC ground truth).

One minus the peak density seems like a reasonable surrogate measure for the human verification threshold. From the datasets annotated in this thesis the \emph{penguins} dataset annotation has the lowest threshold, at approximately $0.15$, and as a result it seems likely the most precise localisation. For the \emph{penguin survey} the threshold is highest at around $0.35$. This makes sense, because counting was emphasised as opposed to precise annotation, so it is natural that the localisation threshold is higher.

@@

For the full amount of training data, $AP_{75}$ is also robust to a small amount of noise or offset, roughly $2\%$ degradation of performance for $4\%$ noise or offset. Human variation of box annotation in \cite{Papadopoulos2017} was reported as a mean \gls{IOU} of 88, which corresponds to the noise case of $\sigma = 4\%$. At that noise level, performance degradation is minimal, even for a strict matching criterion, degrading $AP_{75}$ by around $2\%$. Keep in mind that these datasets were human annotated to begin with, so the noise added is extra noise.

@@

We presented our approach to object detection for the purpose of Verification Based Annotation, where (with a few minor changes) an object detector \cite{Lin2017} can be robust, without a large degree of parameter tweaking, for object detection in a wide variety of domains with a variety of image resolution and object size.

By training with image crops, a higher accuracy is possible using the full image resolution (at the expense of training time, however during annotation, time is plentiful). In each case, using larger crop sizes proved better, more accurate and more stable. Two high-resolution inference methods were were approximately equal, with the tiling approach allowing larger images at the expense of inference time. Incremental training was shown to be practical, with different datasets scaling in accuracy (and in time required for training) differently with the addition of extra images.

An investigation of the impact of annotation localisation noise and systematic error in box annotation show that training with many images is robust to a small amount of localisation error, however, with fewer images the sensitivity is much higher, leading to degradation of performance with even a small amount of noise; this suggests a focus on accurate object localisation is important, especially near the beginning of annotation, for both annotator and tool author.

>>>>

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}
\title{Object detection for Verification Based Annotation}
\author{\IEEEauthorblockN{Oliver Batchelor\IEEEauthorrefmark{1} and
Richard Green\IEEEauthorrefmark{2}}
\IEEEauthorblockA{Department of Computer Science,
University of Canterbury\\
Christchurch, New Zealand\\
Email: \IEEEauthorrefmark{1}oliver.batchelor@canterbury.ac.nz,
\IEEEauthorrefmark{2}richard.green@canterbury.ac.nz}}

@@

\newacronym{NMS}{NMS}{ Non-Maxima Suppression}

@@

In this paper, we examine the properties of an object detector for Verification Based Annotation, where annotation is performed by having machine annotations checked and verified (and corrected, in this work) by a human annotator. We present our approach to object detection, where with a few modifications from standard practice, an object detector \cite{Lin2017} can be robust, without a large degree of parameter tweaking, for training object detection incrementally to annotate a wide variety of small-scale datasets with a variety of image resolution and object size.

Our approach is to use high-resolution images, training on image crops (as opposed to the usual practice of resizing input images to a fixed resolution), and find this method successful, being more accurate and robust than down-scaling. A particular interest is the effect of localisation noise and systematic bias in annotations. We characterise the impact on object detection performance and compare to human levels, we find noise has a large impact, especially with fewer training examples.

@@

In this paper, we examine the properties of an object detector used as part of a \gls{VBA} process. We are most interested in the robustness to varying parameters, such as image size, number and size of objects, and to initialisation. The ability to learn incrementally is key (and in particular, learn from a minimal number of examples).

@@

The object detector is based on a single-shot \gls{CNN} detector called RetinaNet \cite{Lin2017}, an anchor-box based single stage object detector. This detector was selected for its simplicity and efficiency while having close to state-of-the-art accuracy. Some modifications from the method in \cite{Lin2017} and brief motivations for each are:

\begin{itemize}
    \item {\textbf {Sharing}}\par No shared weights between classification sub-networks at different pyramid levels, to make training much faster at the beginning.
    \item {\textbf {Non normalised loss}}\par 
    We do not normalise of the loss function (by the number of anchor matches), to accommodate images with no positive annotations.
    \item {\textbf {High resolution}}\par We train on crops of high-resolution images, but for inference operate on complete images. The motivation is to make the object detection task as easy as possible; larger objects are easier to detect \cite{Lin2014}. 
    \item {\textbf {Cyclical learning rates}}\par We use of cyclical learning rates in training, to accommodate incremental additions to the training set.
    \item {\textbf{Extra decoder layers}}\par We used an extra residual block at each resolution to combine shortcut connections after concatenation with features from lower resolutions.
\end{itemize}

@@

A selection of images of the datasets and typical annotations used in this paper can be seen in Figure~\ref{fig:datasets_all} and a summary of the different variations can be seen in Table~\ref{tab:resolutions}, most of the datasets were annotated by the author, a few annotated by third parties and one was annotated by both. The datasets span a wide variety of image resolutions, and object sizes; the images in some cases are reduced resolution from the captured images, in other cases reducing the resolution would make object detection impossible.

@@

We looked at two different possibilities for performing inference on a full-resolution image using an object detector network trained only on crops of images: (a) pass in the full image using the property that the network is \emph{fully convolutional} and (b) tile images the same size as training crops and collapse the predictions using a combined \gls{NMS}. To facilitate this idea, the box annotations on the edge of images are estimates of the full bounds of the object, and the anchor boxes are not cropped to the edge of the image. In \cite{Duarte2010}, the approach of annotating the estimated full bounding box (the approach we have taken) was compared to annotating just the visible parts (as is more common practice) in a pedestrian detector. Estimating the full bounds was a slight improvement, however, the fusion of both was found to be better than either one.

The first method is to pass in the full image to the neural network, even though it has only been trained on much smaller crops. Object detection networks are flexible and work across a range of input image resolutions. All layers are either convolutions or do not reshape the feature maps (aside from up/down-sampling). Passing in a larger image results in a larger feature map and set of box outputs at each layer of the pyramid. The concern is that the up/down-sampling behaviour is slightly different for input sizes depending on if they are odd or even, relative to powers of two. We found this to be a legitimate concern, though largely negated if the training crop size is set to a power of two or multiple of a power of two.

@@

For this experiment we vary the scale and crop size to test the idea that a high-resolution object detector can be trained on small parts of images, then the whole image used in inference.

@@

Using a larger crop size proved more accurate in all of the four datasets, and using a larger crop was beneficial in all cases, being more accurate and more stable to train. The larger size, however, trains much more slowly. In this case a factor of $3.3$ slower for half scale or half-size crop, larger factors having less speedup due to constants such as loading the images. Despite the relatively low requirement for fast inference (to keep up with a single annotator); this could be used for example for image selection, to find images of most uncertainty for active learning.

@@

When using the annotation tool, image annotations become available incrementally. Here we investigate the difference between training with all examples annotated upfront compared to training with images added to the training set incrementally.

During annotation, the validation set is incrementally built. Here we test against the final validation sets. In future, it would be better to use cross-validation, to make better use of training data, and provide more accurate testing (at the expense of extra training time).

@@

A \gls{VBA} based system by nature of using an object detector trained on noisy inputs, will therefore not produce entirely accurate localised predictions. The hope is that such a system will eventually produce localisations that are \emph{good enough} so that predictions from the object detector can be accepted without taking up valuable annotator time. In this experiment, we aim to quantify how noise degrades performance, and therefore establish some idea of what guideline for the level of precision is required during annotation.

@@

Figure~\ref{fig:density_iou} shows in more detail the distribution of \gls{IOU} overlap for detections which have been modified. The density of transformed detections peaks at around $0.80$--$0.85$ depending on the dataset. It can be reasonably assumed that the real accuracy of the detections, which are used unmodified, would lie between that peak and $1.0$. Human annotation also has some degree of variability (reference \cite{Papadopoulos2017} gave a mean \gls{IOU} overlap of 88 for a box input method with Pascal VOC ground truth).

One minus the peak density seems like a reasonable surrogate measure for the human verification threshold. From the datasets annotated in this thesis the \emph{penguins} dataset annotation has the lowest threshold, at approximately $0.15$, and as a result, it seems likely the most precise localisation. For the \emph{penguin survey} the threshold is highest at around $0.35$. This makes sense because counting was emphasised as opposed to precise annotation, so, the localisation threshold is higher.

@@

For the full amount of training data, $AP_{75}$ is also robust to a small amount of noise or offset, roughly $2\%$ degradation of performance for $4\%$ noise or offset. Human variation of box annotation in \cite{Papadopoulos2017} was reported as a mean \gls{IOU} of 88, which corresponds to the noise case of $\sigma = 4\%$. At that noise level, performance degradation is minimal, even for a strict matching criterion, degrading $AP_{75}$ by around $2\%$. Keep in mind that these datasets were human-annotated to begin with, so the noise added is extra noise.

@@

We presented our approach to object detection for Verification Based Annotation, where (with a few minor changes) an object detector \cite{Lin2017} can be robust, without a large degree of parameter tweaking, for object detection in a wide variety of domains with a variety of image resolution and object size.

By training with image crops, higher accuracy is possible using the full image resolution (at the expense of training time, however during annotation, time is plentiful). In each case, using larger crop sizes proved better, more accurate and more stable. Two high-resolution inference methods were approximately equal, with the tiling approach allowing larger images at the expense of inference time. Incremental training was shown to be practical, with different datasets scaling in accuracy (and in time required for training) differently with the addition of extra images.

An investigation of the impact of annotation localisation noise and systematic error on box annotation shows that training with many images is robust to a small amount of localisation error. However, with fewer images the sensitivity is much higher, leading to degradation of performance with even a small amount of noise; this suggests a focus on accurate object localisation is important, especially near the beginning of annotation, for both annotator and tool author.
